---
title: "EFA Categorical"
output: html_notebook
---
Import Libraries
```{r}
library(foreign) 
library(lavaan)
library(semPlot)
library(tikzDevice)
library(MVN)
library(semTools)
library(performance)
library(psych)
library(tidyverse)
library(multilevel)
library(caTools)
library(car)
library(quantmod)
library(MASS)
library(corrplot)
```


Import Data
```{r}
hz <- read.csv('EFA_cat.csv', header = TRUE)
hz <- subset(hz, select = -c(X))
```

```{r}
#Check for Mulitvariate Normality
mvn(hz, mvn_test = 'mardia')
#Multivariate Normality is violated so we use robust estimator
```


Examine factorability of data
```{r}
correlation <- polychoric(hz)$rho
print(KMO(correlation))

bart.test <- cortest.bartlett(correlation, n = nrow(hz))
print(bart.test)
```
Look at inter item correlations and compare pearson to polychoric. Differences are relatively small and consistent indicating linearity
```{r}
pearson_cor <- cor(hz, use = "pairwise.complete.obs")
poly_cor <- polychoric(hz)$rho

round(pearson_cor, 2)-round(poly_cor, 2)

#This looks reasonably consistent - lets move forward
```
```{r}
#Check Spearman correlation matrix to assess multicollinearity
cor(hz, method = "spearman")
#No correlations above 0.8!
```

Check Initial Correlation
Inspect item-item correlations and the corrected item-total correlations

```{r}
#Get Inter-item Correlations
poly_cor <- polychoric(hz)$rho
round(poly_cor, 2)

#Formula to get polychoric item-total correlations
item_total_cor <- function(R) {
  k <- ncol(R)
  result <- numeric(k)
  
  for (i in 1:k) {
    # indices of all other items
    others <- setdiff(1:k, i)
    
    # sum of correlations between item i and other items
    sum_r_i_others <- sum(R[i, others])
    
    # variance of the total score excluding item i
    var_total_minus_i <- sum(R[others, others])
    
    # corrected item-total correlation
    result[i] <- sum_r_i_others / sqrt(var_total_minus_i)
  }
  
  names(result) <- colnames(R)
  return(result)
}

item_total_cor(poly_cor)
```


Run the initial EFA - Inspect number of factors, item-factor loadings, cross-loadings and parsity of phrasing and factors

```{r}
efa <- fa(hz, nfactors = 2, fm = "wls", rotate = "promax", cor = 'poly')
print(efa)

eigen.values <- eigen(correlation) #Inspect Eigen values
eigen.values$values

scree(hz, pc = FALSE) #Inspect Scree Plot

#Conduct Parallel analysis
paran(hz, iterations = 5000, centile = 0.99, quietly = FALSE, 
    status = TRUE, all = TRUE, graph = TRUE, color = TRUE, 
    col = c("black", "red", "blue"), lty = c(1, 2, 3), lwd = 1, legend = TRUE, 
    file = "", width = 640, height = 640, grdevice = "png", seed = 0)
```

```{r}
library(psych)
library(lavaan)
library(semPlot)

# Run EFA with 2 factors
efa <- fa(hz, nfactors = 2, fm = "wls", rotate = "promax", cor = 'poly')
summary(efa)

# Extract factor loadings
loadings <- as.data.frame(efa$loadings[1:ncol(hz), 1:2])

# Assign each item to the factor with the highest absolute loading
factor_assignments <- apply(loadings, 1, function(x) which.max(abs(x)))
# Load necessary libraries
# 1. Define the file path and resolution
png("EFA_CAT_FINAL.png", width = 8, height = 6, units = "in", res = 300)


psych::fa.diagram(efa, 
                  main = "",          # Remove title for publication
                  simple = TRUE,      # Show only the primary loading for each item
                  cut = 0.3,          # Only show loadings > .30
                  digits = 2, 
                  adj = 1,            # Adjust label position
                  e.size = 0.05,      # Size of the item boxes
                  rsize = 0.15,       # Size of the factor circles
                  side = 1)           # Standard tree orientation

# 3. Save the file
dev.off()
```

```{r}
#Remove Items that violate criteria
hz <- subset(hz, select = -c(X1, X9, X14))

#Recompute and check all
# Run EFA with 2 factors
efa <- fa(hz, nfactors = 2, fm = "wls", rotate = "promax", cor = 'poly')
summary(efa)

# Extract factor loadings
loadings <- as.data.frame(efa$loadings[1:ncol(hz), 1:2])

# Assign each item to the factor with the highest absolute loading
factor_assignments <- apply(loadings, 1, function(x) which.max(abs(x)))
```

Rename the retained variables
```{r}
hz <- hz %>% 
  rename(
    Item1 = X10,
    Item2 = X11,
    Item3 = X12,
    Item4 = X13,
    Item5 = X16,
    Item6 = X18,
    Item7 = X3,
    Item8 = X6,
    Item9 = X7
    )

```

Run the final EFA

```{r}
efa <- fa(hz[, c("Item1" , "Item2" , "Item3" , "Item4" , "Item5" , "Item6" , "Item7" , "Item8" , "Item9")], nfactors = 2, fm = "wls", rotate = "promax", cor = 'poly')

print(efa)
print(efa$communalities)
print(mean(efa$communalities))
print(efa$loadings)
alpha_f1 <- psych::alpha(hz[, c("Item1" , "Item2" , "Item3" , "Item4" , "Item5" , "Item6" , "Item7" , "Item8" , "Item9")])

#Corrected item–total correlations
alpha_f1$item.stats
```


Get the correlations for the final structure
```{r}
#Get Inter-item Correlations
print(min(polychoric(hz[, c("Item1" , "Item2" , "Item3" , "Item4" , "Item5" , "Item6" , "Item7" , "Item8", "Item9")])$rho))
print(mean(polychoric(hz[, c("Item1" , "Item2" , "Item3" , "Item4" , "Item5" , "Item6" , "Item7" , "Item8", "Item9")])$rho))
polychoric(hz[, c("Item1" , "Item2" , "Item3" , "Item4" , "Item5" , "Item6" , "Item7" , "Item8", "Item9")])$rho

item_total_cor(polychoric(hz[, c("Item1" , "Item2" , "Item3" , "Item4" , "Item5" , "Item6" , "Item7" , "Item8", "Item9")])$rho)
```


Get McDonald's Omega
```{r}
# Now compute omega
omega_result <- omega(hz[, c("Item1" , "Item2" , "Item3" , "Item4" , "Item5" , "Item6" , "Item7" , "Item8" , "Item9")], nfactors = 2, fm = 'wls', poly = TRUE)

print(omega_result)
```

Get the factor-factor correlations
```{r}
efa$Phi
```


